Thesis Title

### 2022-08-08 chapters/thesis_outline.tex
Add Grammarly suggestions - thesis overview and introduction

\textbf{Chapter 3} describes the application of both second and third-generation sequencing technologies, also known as next-generation and long-read sequencing, to tap-water samples collected at the University Medical Center Groningen. Our aim was to evaluate the applicability of shotgun metagenomics, but this time in a surveillance setting. Building on the findings from Chapter 2, a hybrid assembly approach was used to increase resolution power. In this sample a new variant of a colistin resistance (mcr) determinants was detected, named mcr-5.4, and through hybrid assembly leveraging both short and long-read sequences, its context was determined, albeit with questionable success. This chapter is included in the following publication: \textit{G. Fleres, N. Couto, L. Schuele, M. A. Chlebowicz, \underline{C. I. Mendes}, L. W. M. van der Sluis, J. W. A. Rossen, A. W Friedrich, S. García-Cobos, Detection of a novel mcr-5.4 gene variant in hospital tap water by shotgun metagenomic sequencing, Journal of Antimicrobial Chemotherapy, Volume 74, Issue 12, December 2019, Pages 3626–3628.  DOI: \url{https://doi.org/10.1093/jac/dkz363.}}

With the lessons learnt in Chapters 2 and 3, we developed in \textbf{Chapter 4} DEN-IM, a one-stop, user-friendly, containerised and reproducible workflow for the analysis of Dengue virus short-read sequencing data from both amplicon and shotgun metagenomics approaches. This takes into particular consideration the dependency on software and database versions used in the metagenomic bioinformatics downstream analysis in the results obtained. Dengue virus represents a public health threat and economic burden in affected countries, with the risk of exposure, increasing, not only driven by travel to endemic regions but also due to the broader dissemination of the mosquito vector, making the burden of dengue very significant. This makes it a particularly relevant target organism for the development of a straightforward workflow for both the identification and characterization of the virus. DEN-IM was designed to perform a comprehensive analysis in order to generate either de novo assemblies or consensus of full viral coding sequences and to identify their serotype and genotype, including the identification of co-infection cases whose prevalence is increasingly found in highly endemic areas. It was developed in Nextflow,  a simple and scalable workflow management system. All tools and dependencies are provided in Docker containerised images. All these steps ensure reproducibility and transparency of the bioinformatic process. This chapter is included in the following publication: \textit{\underline{C. I. Mendes}\footnote[1]{These authors contributed equally to this work.}, E. Lizarazo$^*$ , M. P. Machado, D. N. Silva, A. Tami, M. Ramirez, N. Couto, J. W. A. Rossen, J. A. Carriço, DEN-IM: dengue virus genotyping from amplicon and shotgun metagenomic sequencing. Microbial Genomics, Volume 6, Issue 3, March 2020. DOI: \url{https://doi.org/10.1099/mgen.0.000328.}}

A key process in metagenomic data analysis is the de novo assembly of raw sequence data since it allows recovering contigs representing the replicons present in the sample, be it genomes, plasmids, or bacteriophages, from a pool of mixed raw reads. \textbf{Chapter 5} employs the same core principles as in Chapter 4, describing a one-stop, user-friendly, containerised, and reproducible workflow, named LMAS, to assess the performance of de novo assembly algorithms for the assembly of second-generation metagenomic sequencing data. The LMAS workflow, which allows users to evaluate performance given a known standard community was implemented in Nextflow, ensuring the transparency and reproducibility of the results obtained. Similarly to Chapter 4, the use of Docker containers provides additional flexibility. The results are presented in an interactive HTML report where global and reference specific performance metrics can be explored. Currently, 12 de novo assemblers are implemented in LMAS, with the possibility of expansion as novel algorithms are developed.

@@

\textbf{Chapter 8} showcases an effort to raise standards on the development and distribution of code for bioinformatic analysis. For this, seven recommendations are presented that help researchers implement software testing in microbial bioinformatics. We propose collaborative software testing as an opportunity to continuously engage software users, developers, and students to unify scientific work across domains. As automated software testing remains underused in scientific software, our set of recommendations not only ensures that appropriate effort can be invested in producing high quality and robust software, but also increases engagement in its sustainability. This chapter is included in the following publication: \textit{B. C. L. van der Putten$^*$, \underline{C. I. Mendes}\footnote[1]{These authors contributed equally to this work.}, B. M. Talbot, J. de Korne-Elenbaas, R. Mamede, P. Vila-Cerqueira, L. P. Coelho, C. A. Gulvik, L. S. Katz, The Asm Ngs Hackathon Participants, Software testing in microbial bioinformatics: a call to action. Microbial Genomics, Volume 8, Issue 3. DOI: \url{https://doi.org/10.1099/mgen.0.000790.}}

>>>>

\textbf{Chapter 3} describes the application of both second and third-generation sequencing technologies, also known as next-generation and long-read sequencing, to eight tap-water samples collected at the University Medical Center Groningen. Our aim was to evaluate the applicability of shotgun metagenomics, but this time in a surveillance setting. Building on the findings from Chapter 2, a hybrid assembly approach was used to increase resolution power. In this sample a new variant of a colistin resistance (mcr) determinant was detected, named mcr-5.4, and through hybrid assembly leveraging both short and long-read sequences, its context was determined, albeit with questionable success. This chapter is included in the following publication: \textit{G. Fleres, N. Couto, L. Schuele, M. A. Chlebowicz, \underline{C. I. Mendes}, L. W. M. van der Sluis, J. W. A. Rossen, A. W Friedrich, S. García-Cobos, Detection of a novel mcr-5.4 gene variant in hospital tap water by shotgun metagenomic sequencing, Journal of Antimicrobial Chemotherapy, Volume 74, Issue 12, December 2019, Pages 3626–3628.  DOI: \url{https://doi.org/10.1093/jac/dkz363.}}

With the lessons learnt in Chapters 2 and 3, we developed in \textbf{Chapter 4} DEN-IM, a one-stop, user-friendly, containerised and reproducible workflow for the analysis of Dengue virus short-read sequencing data from both amplicon and shotgun metagenomics approaches. This takes into particular consideration the dependency on software and database versions used in the metagenomic bioinformatics downstream analysis in the results obtained. Dengue virus represents a public health threat and economic burden in affected countries, with the risk of exposure, increasing, not only driven by travel to endemic regions but also due to the broader dissemination of the mosquito vector, making the burden of dengue very significant. This makes it a particularly relevant target organism for the development of a straightforward workflow for both the identification and characterization of the virus. DEN-IM was designed to perform a comprehensive analysis in order to generate either de novo assemblies or consensus of full viral coding sequences and to identify their serotype and genotype, including the identification of co-infection cases whose prevalence is increasingly found in highly endemic areas. It was developed in Nextflow,  a simple and scalable workflow management system. All tools and dependencies are provided in Docker containerised images. All these steps ensure reproducibility and transparency of the bioinformatic process. This chapter is included in the following publication: \textit{\underline{C. I. Mendes}\footnote[1]{These authors contributed equally to this work.}, E. Lizarazo$^*$, M. P. Machado, D. N. Silva, A. Tami, M. Ramirez, N. Couto, J. W. A. Rossen, J. A. Carriço, DEN-IM: dengue virus genotyping from amplicon and shotgun metagenomic sequencing. Microbial Genomics, Volume 6, Issue 3, March 2020. DOI: \url{https://doi.org/10.1099/mgen.0.000328.}}

A key process in metagenomic data analysis is the de novo assembly of raw sequence data since it allows recovering contigs representing the replicons present in the sample, be it genomes, plasmids, or bacteriophages, from a pool of mixed raw reads. \textbf{Chapter 5} employs the same core principles as in Chapter 4, describing a one-stop, user-friendly, containerised, and reproducible workflow, named LMAS, to assess the performance of de novo assembly algorithms for the assembly of second-generation metagenomic sequencing data. The LMAS workflow, which allows users to evaluate performance given a known standard community was implemented in Nextflow, ensuring the transparency and reproducibility of the results obtained. Similarly to Chapter 4, the use of Docker containers provides additional flexibility. The results are presented in an interactive HTML report where global and reference-specific performance metrics can be explored. Currently, 12 de novo assemblers are implemented in LMAS, with the possibility of expansion as novel algorithms are developed.

@@

\textbf{Chapter 8} showcases an effort to raise standards on the development and distribution of code for bioinformatic analysis. For this, seven recommendations are presented that help researchers implement software testing in microbial bioinformatics. We propose collaborative software testing as an opportunity to continuously engage software users, developers, and students to unify scientific work across domains. As automated software testing remains underused in scientific software, our set of recommendations not only ensures that appropriate effort can be invested in producing high-quality and robust software but also increases engagement in its sustainability. This chapter is included in the following publication: \textit{B. C. L. van der Putten$^*$, \underline{C. I. Mendes}\footnote[1]{These authors contributed equally to this work.}, B. M. Talbot, J. de Korne-Elenbaas, R. Mamede, P. Vila-Cerqueira, L. P. Coelho, C. A. Gulvik, L. S. Katz, The Asm Ngs Hackathon Participants, Software testing in microbial bioinformatics: a call to action. Microbial Genomics, Volume 8, Issue 3. DOI: \url{https://doi.org/10.1099/mgen.0.000790.}}
